%% Settings for single-side (simplex) printing
\documentclass[12pt,a4paper]{report}
\setlength\textwidth{145mm}
\setlength\textheight{247mm}
\setlength\oddsidemargin{15mm}
\setlength\evensidemargin{15mm}
\setlength\topmargin{0mm}
\setlength\headsep{0mm}
\setlength\headheight{0mm}
\let\openright=\clearpage

\parskip = \baselineskip
\divide \parskip by 4

\usepackage[utf8]{inputenc}
\usepackage{amsmath}        % extensions for typesetting of math
\usepackage{amsfonts}       % math fonts
\usepackage{amsthm}         % theorems, definitions, etc.
\usepackage{bbding}         % various symbols (squares, asterisks, scissors, ...)
\usepackage{bm}             % boldface symbols (\bm)
\usepackage{graphicx}       % embedding of pictures
\usepackage{fancyvrb}       % improved verbatim environment
\usepackage{natbib}         % citation style AUTHOR (YEAR), or AUTHOR [NUMBER]
\usepackage[nottoc]{tocbibind} % makes sure that bibliography and the lists
			    % of figures/tables are included in the table
			    % of contents
\usepackage{dcolumn}        % improved alignment of table columns
\usepackage{booktabs}       % improved horizontal lines in tables
\usepackage{paralist}       % improved enumerate and itemize
\usepackage[usenames]{xcolor}  % typesetting in color
\usepackage{listings}
\usepackage[T1]{fontenc}
\usepackage{enumitem}
\usepackage{tabularx}
\usepackage[intoc,refpage]{nomencl}

\lstset{ %
  breakatwhitespace=true,
  breaklines=true,
  keepspaces=true,
  columns=fullflexible,
  basicstyle=\small\ttfamily,
  numbers=none,
  frame=single,
  xleftmargin=\parindent,
  xrightmargin=\parindent,
  aboveskip=\bigskipamount,
  belowskip=\bigskipamount,
  framesep=6pt,
  showstringspaces=false,
  showtabs=false,
  tabsize=2
}

%%% Basic information on the thesis
\def\ThesisTitle{Multicast routing}
\def\ThesisAuthor{Ondřej Hlavatý}
\def\YearSubmitted{2016}
\def\Department{Department of Applied Mathematics}
\def\DeptType{Department}
\def\Supervisor{Mgr. Martin Mareš, Ph.D.}
\let\SupervisorsDepartment\Department
\def\StudyProgramme{Computer Science}
\def\StudyBranch{IOI}

% An optional dedication: you can thank whomever you wish (your supervisor,
% consultant, a person who lent the software, etc.)
\def\Dedication{%
  I would like to thank Ondřej `Santiago' Zajíček from the CZ.NIC association
  for giving me directions on the topic and quickly cutting the dead ends, and
  my supervisor Martin Mareš for many interesting discussions.
}

% Abstract (recommended length around 80-200 words; this is not a copy of your thesis assignment!)
\def\Abstract{%
  We implement multicast routing in the BIRD routing daemon.
  At first, we review basic principles of IP multicast and different
  modes of the PIM multicast routing protocol.
  Then we outline the architecture of the BIRD and discuss its aspects
  related to our goal. We present our solution based on adding IGMP,
  bidirectional PIM, and kernel interface as three separate BIRD protocols.
  Finally, we describe our framework for testing network software
  in complex virtual networks.
}

\def\CzAbstract{
  V práci implementujeme podporu multicastového směrování do směrovacího démonu
  BIRD. Nejprve projdeme základní principy multicastu v IP a různé režimy
  multicastového směrovacího protokolu PIM. Následně rozebíráme aspekty démonu
  BIRD související s naším cílem. Představujeme řešení spočívající
  v implementaci třech samostatných BIRD protokolů: IGMP, PIM a multicastového
  rozhraní k jádru OS. Nakonec popisujeme naše nástroje pro testování síťového
  software jako je BIRD v komplexních virtuálních sítích.
}

\def\Keywords{%
  {multicast routing}, {BIRD}, {PIM}, {networking}
}

\usepackage[pdftex,unicode]{hyperref}   % Must follow all other packages
\hypersetup{breaklinks=true}
\hypersetup{pdftitle={\ThesisTitle}}
\hypersetup{pdfauthor={\ThesisAuthor}}
\hypersetup{pdfkeywords=\Keywords}
\hypersetup{urlcolor=blue}

\include{macros}

\makenomenclature

% Title page and various mandatory informational pages
\begin{document}
\include{title}

\tableofcontents

\include{preface}
\include{chap01} % Multicast in IPv4
\include{chap02} % PIM
\include{chap03} % The BIRD
\include{chap04} % Multicast in the BIRD -- implementation
\include{chap05} % Multicast in the BIRD -- usage
\include{chap06} % Testing
\include{epilog}
\include{bibliography}

\renewcommand\nomname{List of Abbreviations}
\printnomenclature[2.5cm]

\openright
\end{document}
